High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-04-03 notebooks/intro.qmd
uh

temp = 0.5
p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
plot(p1, p2, size=(800,320))
```

@@

plts = []
for (i,target) ∈ enumerate(counterfactual_data.y_levels)
    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
    scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=i,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

# Generators:
generators = Dict(
    ordered_names[1] => generator,
    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0set_size_penalty),
    ordered_names[3] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0distance_from_energy),
    ordered_names[4] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0distance_from_targets),
)

@@

# Plots:
plts = []
for name ∈ ordered_names
    ce = counterfactuals[name]
    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
    if :energy_sampler ∈ collect(keys(ce.params))
        Xgen = ce.params[:energy_sampler].buffer
        scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    end
    plts = vcat(plts..., plt)
end
_n = length(generators)
plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
```

### Benchmark

# Data:
datasets = Dict(
    :linearly_separable => load_linearly_separable(),
    :overlapping => load_overlapping(),
    :moons => load_moons(),
    :circles => load_circles(),
)

# Models:
models = Dict(
    :MLP => M,
)
```

```{julia}
#| output: false
using CounterfactualExplanations.Evaluation: benchmark
bmks = []
for (dataname, dataset) in datasets
    bmk = benchmark(dataset; models=models, generators=generators, suppress_training=true, dataname=dataname)
    push!(bmks, bmk)
end
bmk = reduce(vcat, bmks)
```

@@

temp = 0.5
p1 = contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp, clim=nothing)
p2 = contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=-1, temp=temp, clim=nothing, loss_matrix=ones(4,4))
plot(p1, p2, size=(800,320))
```

@@

plts = []
for target ∈ counterfactual_data.y_levels
    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = plot(M, counterfactual_data; target=target, zoom=-0.5,cbar=false)
    scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

# Plots:
plts = []
for name ∈ ordered_names
    ce = counterfactuals[name]
    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
    if :energy_sampler ∈ collect(keys(ce.params))
        Xgen = ce.params[:energy_sampler].buffer
        scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    end
    plts = vcat(plts..., plt)
end
_n = length(generators)
plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
```

>>>>

temp = 0.5
p1 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=0, temp=temp)
p2 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=0, temp=temp, clim=nothing, loss_matrix=ones(2,2))
Plots.plot(p1, p2, size=(800,320))
```

@@

plts = []
for (i,target) ∈ enumerate(counterfactual_data.y_levels)
    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=i,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

# Generators:
generators = Dict(
    ordered_names[1] => generator |> gen -> @objective(gen, _ + 1.0distance),
    ordered_names[2] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0set_size_penalty),
    ordered_names[3] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0distance_from_energy),
    ordered_names[4] => deepcopy(generator) |> gen -> @objective(gen, _ + 0.1distance + 10.0distance_from_targets),
)

@@

# Plots:
plts = []
for name ∈ ordered_names
    ce = counterfactuals[name]
    plt = Plots.plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
    if :energy_sampler ∈ collect(keys(ce.params))
        Xgen = ce.params[:energy_sampler].buffer
        Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    end
    plts = vcat(plts..., plt)
end
_n = length(generators)
Plots.plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
```

@@

temp = 0.5
p1 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_set_loss=true, zoom=-1, temp=temp, clim=nothing)
p2 = Plots.contourf(mach.model, mach.fitresult, X, y; plot_classification_loss=true, zoom=-1, temp=temp, clim=nothing, loss_matrix=ones(4,4))
Plots.plot(p1, p2, size=(800,320))
```

@@

plts = []
for target ∈ counterfactual_data.y_levels
    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-0.5,cbar=false)
    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

# Plots:
plts = []
for name ∈ ordered_names
    ce = counterfactuals[name]
    plt = Plots.plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=-0.5)
    if :energy_sampler ∈ collect(keys(ce.params))
        Xgen = ce.params[:energy_sampler].buffer
        Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    end
    plts = vcat(plts..., plt)
end
_n = length(generators)
Plots.plot(plts..., size=(_n * img_height,1.05*img_height), layout=(1,_n))
```

## Benchmarks

# Data:
datasets = Dict(
    :linearly_separable => load_linearly_separable(),
    :overlapping => load_overlapping(),
    :moons => load_moons(),
    :circles => load_circles(),
    :multi_class => load_multi_class(),
)

# Untrained Models:
models = Dict(
    :cov75 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
    :cov80 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
    :cov90 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
    :cov99 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
)
```

```{julia}
#| output: false
using CounterfactualExplanations.Evaluation: benchmark
bmks = []
measures = [
    CounterfactualExplanations.distance,
    CCE.distance_from_energy,
    CCE.distance_from_targets
]
for (dataname, dataset) in datasets
    bmk = benchmark(
        dataset; 
        models=deepcopy(models), 
        generators=generators, 
        measure=measures,
        suppress_training=false, dataname=dataname,
        n_individuals=10
    )
    push!(bmks, bmk)
end
bmk = reduce(vcat, bmks)
```

```{julia}
@chain bmk() begin
    @group_by(model, dataname)
    @select(model, dataname, ce)
    @mutate(performace = CounterfactualExplanations.model_evaluation(ce.M, ce.data))
    @summarize(model=unique(model), dataname=unique(dataname), performace=unique(performance))
    @ungroup
end
```

```{julia}
#| output: true
#| echo: false
#| label: fig-benchmark
#| fig-cap: "Benchmark results for the different generators."
#| fig-subcap: 
#|  - "Circles."
#|  - "Linearly Separable."
#|  - "Moons."
#|  - "Multi-class."
#|  - "Overlapping."

df = bmk()
for dataname ∈ sort(unique(df.dataname))
    df_ = df[df.dataname .== dataname, :]
    plt = data(df_) * visual(BoxPlot) * 
        mapping(:generator, :value, row=:variable, col=:model, color=:generator)
    plt = draw(
        plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=200, height=180), 
        # facet=(; linkyaxes=:none)
    )   
    display(plt)
end
```
